One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2020-02-21 paper-v2.tex
end of Janine's comments

\begin{abstract}
Distance sampling methods aim to estimate the size and spatial distribution of populations of wild animals.  Here we present an analysis of point transect distance sampling data collected on a tropical forest bird.  We consider an entire workflow, from model conception, inference procedure, model evaluation to communication of results, at each stage discussing the challenges involved and our strategies to address them.

New developments inlcude an inference method for one-stage distance sampling models within a Bayesian framework based on iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations \citep{rue_approximate_2009}.  This approach simultaneously estimates the detectability and spatial distribution of populations of wild animals within a joint model.  We take a point process perspective, modelling the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields implemented via a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.  The Bayesian setting allows sampling from the joint posterior of all model parameters and so the resulting model outputs naturally incorporate uncertainty in the observation model estimates.  We discuss potential issues with common ways to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels.  We apply the methods of excursion sets and excursion functions \citep{bolin_excursion_2015} as an example of this approach.
\end{abstract}

%\begin{keyword}
%Distance sampling \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation \sep Generalized additive model
%\end{keyword}
\end{frontmatter}

@@

The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}. Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.  The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century [[Rick - citation possible here?]].  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.  This information is critical when decisions about conservation strategies are made.  However, answering these questions presents many statistical challenges.

@@

Thirdly, a key aim of the analysis is to predict animal density at un-sampled locations.  If density is modelled using spatial covariates the usual approach is to assume the relationship between the animals and spatial covariates is the same in the un-sampled locations as it is at the sampled locations.  Predictions can then be based on estimating this relationship using the observed data.  However, we usually have reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g. the 'sociality' of the species or other behavioural characteristics).  For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE: Ask Rick if poss]].  From a statistical perspective, this suggests the inclusion of spatially-structured random effects in the model, which necessitates the use of appropriate statistical software capable of fitting such models.

@@

@@

Using spatially-structured random effects can be computationally expensive, often resulting in dense prior precision matrices which typically have $\mathcal{O}(n^3)$ cost to factorise for an effect with $n$ parameters - sometimes referred to as the `big $n$ problem'.  A key development in this area is the use of Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).  This representation is constructed via a stochastic partial differential equation solved using finite element methods \citep{lindgren_explicit_2011}.

@@

The results presented above are consistent with those resulting from a two-stage model fitted to binned count data (see Supplementary material).  Our presentation of the results is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided there and it is common to see maps of point estimates without any accompanying communication of uncertainty.  Our intention in this section is to highlight limitations with these maps, including those presented in Section \ref{results}, and suggest ways to address these limitations.

@@

Another natural extension is to consider more complicated observation processes.  For example, we did not use any explanatory covariates for the detection function. Any additional parameters to non-linear model components will become part of the outer-optimisation step of iterated INLA.

We note also that our presentation of abundance estimates differs from typical presentations.  It is common to report a point estimate of the expected abundance and uncertainty in this point estimate.  However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.  We find the distinction between expected and realised abundance to be a useful one. The credible intervals for the posterior expected abundance will be narrower than the credible intervals for the posterior realised abundance.  The Bayesian perspective is useful here as well as for the uncertainty in the detection model estimates.  Since outputs based on sampling from the joint posterior of all model parameters, including the detection model, the Bayesian framework means the resulting model outputs we present here naturally average over the uncertainty in the observation process.  [[Not really sure if this stuff is worth mentioning or if can be said in a better way]]

Our use of \texttt{inlabru} and \texttt{R-INLA} for inference means we can take advantage of the sparsity of the GMRF precision matrix, a key advantage of the SPDE approach.  A dense covariance matrix of dimension $n \times n$ typically has an $\mathcal{O}(n^3)$ cost to invert.  In contrast to this \texttt{R-INLA} makes use of sparse matrix algorithms to invert a sparse 2-dimensional precision matrix at a typical cost of $\mathcal{O}(n^{3/2})$ \citep{lindgren_explicit_2011}.  This computational efficienty, in addition to the flexibility of the mesh triangulation, can make the SPDE an attractive choice for implementing high-resolution spatial effects in complex domains.  An additional benefit, as described in detail in \cite{yuan_point_2017}, is that the SPDE can also be implemented on the sphere, making analysis of global datasets computationally feasible.

In our analysis we only briefly touch upon the problem of communicating uncertainty in maps and highlight the excursions methods as a new and positive addition to this area.  Our example threshold was chosen arbitrarily and it is not obvious to us that there will always be an easy or simple choice of thresholds and levels of uncertainty for a given analysis.  This requires the input of relevant domain expertise and statistical knowledge so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.  The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.  Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions.  In the above we chose to use excursion sets, excursion functions and the posterior of realised abundance as a demonstration of this.  It is not always the case that a map of predicted mean abundance will be the best output for a species distribution model.

The point process perspective with spatial SPDE effect can seem fundamentally different to a two-stage GAM approach.  However, we think that it can be useful to highlight the similarities in approach.  It can be tempting to think that because with the point process model does not require us to bin the data into counts that we can avoid some loss of information.  However, we also had to make the same assumption of constant intensity within each transect that is common in more traditional distance sampling approaches.  This means the exactly location of the points is not used to inform the posterior intensity.  The assumption of constant intensity within each point transect is essentially the point process version of binning point data into counts.

The SPDE effect can also seem fundamentally different to, say, a spline-based smoother.  However, there is a fundamental connection between SPDEs and optimal smoothing splines.  The optimal smoothing spline is a lminimum variance unbiased linear estimator of the posterior mean of an SPDE \citep{kimeldorf_spline_2002}.  This means a smoothing spline can be interpreted as an SPDE and vice-versa.  Depending on the methods a user is most familiar with this can be a useful perspective.  \cite{yue_bayesian_2014} show how to implement common spline penalties in an SPDE framework and \cite{miller_understanding_2019} offer a complementary explanation to interpret and fit SPDE models within standard penalised regression splines software.

Although we use a Bayesian approach, maximum-likelihood frameworks could also have been used for a one-stage analysis.  
Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analysis.

Discuss selectivity, saturated responses?? Disease modelling?

\clearpage
\bibliographystyle{plainnat}
\bibliography{paper}

>>>>

\begin{abstract}
Distance sampling methods aim to estimate the size and spatial distribution of populations of wild animals.  Here we present an analysis of point transect distance sampling data collected on a tropical forest bird.  We consider the full workflow from model specification, inference procedure, model evaluation and communication of results, at each stage discussing the statistical challenges involved and our strategies to address them.  We take a novel point process perspective on point transect data, viewing observations as a thinned log-Gaussian Cox process with a spatially structured random effect specified via a computationally efficient Gaussian Markov random field approximation to Gaussian random field \citep{lindgren_explicit_2011}.  One-stage inference is achieved within a Bayesian framework by using an approach based on iterated fitting of latent Gaussian models using integrated nested Laplace approximations \citep{rue_approximate_2009} to account for non-linear model components, simultaneously estimating detectability and spatial distribution.  Model outputs are based on samples from the joint posterior of all model parameters and so naturally incorporate uncertainty due to detectability, a key advantage of the one-stage Bayesian approach.  We discuss limitations of methods to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels by using excursions methods \citep{bolin_excursion_2015} to avoid these issues. The methods presented are general, flexible and applicable in a wide range of areas in spatial statistics beyond just species distribution mapping.
\bigskip
\end{abstract}

\begin{keyword}
Distance sampling \sep Species distribution modelling \sep Estimating abundance \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation
\bigskip
\end{keyword}
\end{frontmatter}

@@

The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}. Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.  The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century [[CITE]].  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.  This information is critical when decisions about conservation strategies are made.  However, answering these questions presents many statistical challenges.

@@

Thirdly, a key aim of the analysis is to predict animal density at un-sampled locations.  If density is modelled using spatial covariates the usual approach is to assume the relationship between the animals and spatial covariates is the same in the un-sampled locations as it is at the sampled locations.  Predictions can then be based on estimating this relationship using the observed data.  However, we usually have reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g. the 'sociality' of the species or other behavioural characteristics).  For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE]].  From a statistical perspective, this suggests the inclusion of spatially-structured random effects in the model, which necessitates the use of appropriate statistical software capable of fitting such models.

@@

The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.  Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions and decisions.  In this paper we provide an example based on a scenario where the most important aim of the analysis is identification of areas where animal density is above a certain threshold.  It is not always the case that the most appropriate output of a species distribution model is a map of the predicted mean density.

@@

Using spatially-structured random effects can be computationally expensive, often resulting in dense prior precision matrices which typically have $\mathcal{O}(n^3)$ cost to factorise for an effect with $n$ parameters.  A key development in this area is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs) and the \texttt{R-INLA} software that makes use of the sparsity to factorise the precision matrix at a cost of $\mathcal{O}(n^{3/2})$.  This representation is constructed via a stochastic partial differential equation solved using finite element methods \citep{lindgren_explicit_2011}.

@@

The results above are consistent with those resulting from a two-stage model fitted to binned count data (see Supplementary material).  Our presentation of the results is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided there and it is common to see maps of point estimates without any accompanying communication of uncertainty.  It is also common to report a point estimate of the expected abundance along with uncertainty in this point estimate.  However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.  For this reason we chose to show the full abundance posterior in \autoref{fig:realized-abundance-posterior}.

Our intention in this section is to highlight limitations with mapped summaries of model predictions, including those presented in Section \ref{results}, and suggest ways to address these limitations.  We note that all model outputs are based on sampling from the joint posterior of all model parameters which naturally average over the uncertainty in the observation process, a key advantage of the one-stage approach.

@@

Another natural extension is to consider more complicated observation processes.  There is the potential to include explanatory covariates on detection function parameters. Any additional parameters to the non-linear model components will be estimated using the outer-optimisation step of iterated INLA. [[What can I say about this?  Speculate on how many parameters would work?]]

The iterated INLA inference procedure is a general approach that could be applied to other non-linear model components that arise in the area of spatial ecology.  We note two promising areas in which our flexible inference procedure may be useful:

\begin{itemize}
	\item \textbf{Selectivity analysis} In surveys of the marine environment, the size, length and weight of fish affects the likelihood of being caught in a trawl net.  Thus any analysis of these data must account for this.  The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}, which could be estimated jointly with the spatial distribution within our inference framework.  
	\item \textbf{Functional responses} In the area of understanding species-habitat or species-species interactions, the concept of a `functional response' captures the idea that the abundance of a resource or risk will affect the response to it \citep{holling_some_1959}.  These varying responses can be described mathematically using parametric equations that have the desired properties, e.g. a horizontal asymptote so that beyond a certain threshold no extra amount of a resource will affect the response to it.  These types of responses are typically not considered in species distribution models but our framework allows them to be estimated alongside the more common linear or quadratic fixed effects. This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may be non-linear.

\end{itemize}
We note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analysis

In our analysis we only briefly touch upon the problem of communicating uncertainty in maps and highlight the excursions methods as a new and positive addition to this area.  In our example, the threshold was chosen to illustrate the method but we believe that discussions with relevant stakeholders will lead to agreement on a (possibly large) set of thresholds that are relevant given the aims and context of the analysis.  This requires the input of both the relevant domain experts and statisticians so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.  We also note the potential for excursions to be used to define the notion of occupancy (and thus map occupied areas, with a given probability of error).

We end with a note on the similarity between what we have done and existing two stage approach.  The one-stage thinned point process perspective can seem fundamentally different to a two-stage GAM approach fitted to count data.  It can be tempting to think that, because the point process model does not require us to bin the data into counts, we can avoid some loss of information.  However, we also made the same assumption of constant intensity within each transect as in more traditional distance sampling approaches.  This leads to the same loss of information since we do not use the exact location of points within each transect, except to account for detectability.

In summary, we have presented a novel framework for analysing point transect distance sampling data that introduces new approaches to model specification, inference procedure and communication of results.  These developments will be of interest to those with similar data structures in spatial ecology as well as researchers in other areas who we hope will explore the full flexibility and generality of these developments in their own work.

% spp has potential to provide general framework? [Janine comment]
% Mention modelling on sphere is possible without projecting on to R^2?

\clearpage
\bibliographystyle{humannat}
\bibliography{paper}
